---
output: github_document
---

<!-- README.md is generated from the source: README.Rmd -->

# distcomp

[![Travis build status](https://travis-ci.org/bnaras/distcomp.svg?branch=master)](https://travis-ci.org/bnaras/distcomp)
[![CRAN_Status_Badge](http://www.r-pkg.org/badges/version/distcomp)](https://cran.r-project.org/package=distcomp)
[![Coverage
Status](https://img.shields.io/codecov/c/github/bnaras/distcomp/master.svg)](https://codecov.io/github/bnaras/distcomp?branch=master)
[![](https://cranlogs.r-pkg.org/badges/distcomp)](https://cran.r-project.org/package=distcomp)


This package is described in detail in the paper [_Software for Distributed Computation on Medical Databases: A Demonstration Project._](https://doi.org/10.18637/jss.v077.i13)

## Installation

Install this package the usual way in R or via:

```{r, eval = FALSE}
library(devtools)
install_github("bnaras/distcomp")
```

Then, you will find a document that describes several examples
installed under the R library tree. For example:

```{r, eval = FALSE}
list.files(system.file("doc", package = "distcomp"))
list.files(system.file("doc_src", package = "distcomp"))
```

The examples described in the reference below are available as
follows:

```{r, eval = FALSE}
list.files(system.file("ex", package = "distcomp"))
```

Use of this package requires some configuration. In particular, to run
the examples on a local machine where a single `opencpu` server will
be emulating several sites, a suitable R profile needs to be set
up. That profile will be something along the lines of

```{r, eval=FALSE}
library(distcomp)
distcompSetup(workspace = "full_path_to_workspace_directory",
              ssl_verifyhost = 0L, ssl_verifypeer = 0L)
```
where the workspace is a directory that the `opencpu` server can
serialize objects to. On Unix or Mac, the above can be inserted into
an `.Rprofile` file, but on Windows, we find that the `Rprofile.site`
file needs to contain the above lines.

The effect of this is that _every R_ process (including the `opencpu`
process) has access to the `distcomp` library and the workspace.

## Prototyping New Computations

Refer to the vignette in the package for some tips on developing new
distributed computations. 

## References

Balasubramanian Narasimhan and Daniel Rubin and Samuel Gross and
Marina Bendersky and Philip Lavori. Software for Distributed
Computation on Medical Databases: A Demonstration Project. Journal of
Statistical Software, Volume 77, Issue 13, (2017). [DOI](https://dx.doi.org/10.18637/jss.v077.i13)
